Models

### 2021-07-29 analysis/05_model_metrics.Rmd
Mapping approach instead of tidymodels

pins::pin(model_data, board = "github")
```
Add human footprint data
```{r}
hfp_data <- pins::pin_get("hfp-data", board = "github")

model_data_hfp <- model_data %>% 
  #dplyr::select(rarefyID, year) %>% 
  #distinct() %>%
  mutate(hfp_year = case_when(
    year < 2001 ~ 1993,
    TRUE ~ 2009
  ), 
  hfp_rarefyID = str_replace_all(rarefyID, pattern = "_bird", ""), 
  hfp_rarefyID = str_replace_all(hfp_rarefyID, pattern = "_mamm", "")) %>%
  left_join(hfp_data %>% dplyr::select(-STUDY_ID) %>% mutate(year = as.numeric(year)), by = c("hfp_year" = "year", "hfp_rarefyID" = "rarefyID")) %>%
  dplyr::select(-hfp_year)

@@

```{r}
SES_rand_slope <- model_data %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(#year_center = map(data, ~scale(.x$year)),
         model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
           #                       data = .x)),
           # test = map2(model, null_model, anova),
           # test_res = map(test, tidy),
           coef = map(model, tidy),
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

#check that they all used the same number of observations
map(SES_rand_slope$model, nobs)

#these two metrics are cranky and want their own year scaling to converge

#Jaccard Base
jacc_base_mod <-  model_data %>%
  filter(metric == ("Jaccard_base")) %>%
  mutate(year_scaled = scale(year)) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           coef = map(model, tidy),
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

# S model
s_mod <- model_data %>%
  filter(metric %in%  c("S")) %>%
  #it wants its own scaling to converge for some reason
  mutate(year_scaled = scale(year)) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)

#save coefficients and models
usethis::use_data(metric_model_table)
pins::pin(metric_model_table, board = "github")
```

@@

#### Analizing slopes
Is slope a function of timeseries characteristics?
```{r}
study_slope_data <- study_slopes %>%
  select(study_id, metric, estimate) %>%
  left_join(meta %>% 
              select(-rarefyid) %>% 
              distinct() %>%
              mutate(study_id = as.character(study_id))) %>%
  left_join(model_data %>% select(study_id, startyear, duration) %>% distinct())

#taxa model
taxa_slope <- study_slope_data %>%
  mutate(taxa = as.factor(taxa)) %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~glm(estimate ~ taxa, 
                                 data = .x)),
           coef = map(model, tidy),
           ) %>% 
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  mutate(covariate = "taxa")

climate_slope <- study_slope_data %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(#year_center = map(data, ~scale(.x$year)),
         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
         model = map(data, ~glm(estimate ~ climate, 
                                 data = .x)),
           coef = map(model, tidy),
           ) %>% 
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  mutate(covariate = "climate")

realm_slope <- study_slope_data %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(#year_center = map(data, ~scale(.x$year)),
         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
         model = map(data, ~glm(estimate ~ realm, 
                                 data = .x)),
           coef = map(model, tidy),
           ) %>% 
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  mutate(covariate = "realm")

>>>>

pins::pin(model_data, board = "github")
```
Add human footprint data
```{r}
# hfp_data <- pins::pin_get("hfp-data", board = "github")
# 
# model_data_hfp <- model_data %>% 
#   #dplyr::select(rarefyID, year) %>% 
#   #distinct() %>%
#   mutate(hfp_year = case_when(
#     year < 2001 ~ 1993,
#     TRUE ~ 2009
#   ), 
#   hfp_rarefyID = str_replace_all(rarefyID, pattern = "_bird", ""), 
#   hfp_rarefyID = str_replace_all(hfp_rarefyID, pattern = "_mamm", "")) %>%
#   left_join(hfp_data %>% dplyr::select(-STUDY_ID) %>% mutate(year = as.numeric(year)), by = c("hfp_year" = "year", "hfp_rarefyID" = "rarefyID")) %>%
#   dplyr::select(-hfp_year)

@@

Function approach to model
```{r}
get_model <- function(metric_name, formula, model_id, data, ...){
  print(c(metric_name, model_id))
  qlmer <- quietly(lmer)

  fit <- data %>%
    filter(metric == metric_name) %>%
    mutate(year_scaled = scale(year)) %>%
    qlmer(formula = formula, data = ., ...)

  tidy(fit$result) %>%
   mutate(metric = metric_name, model = list(fit), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
}

get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
```

Create dataframe of models to fit
```{r}
raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base"), 
                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
                         taxa = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
                         realm = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
                         climate = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

log_value_mods <- tibble(metric_name = c("S"), 
                         base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
                         taxa = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
                         realm = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
                         climate = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
  select(metric_name, formula, model_id)
```

Fit models
```{r}
#Fit models with base settings
mod_results <- pmap_dfr(mods, get_model, data = model_data)

#get the models that had convergence warnings
bad_models <- mod_results %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(mods, by = c("metric" = "metric_name", "model_id"))

#Turn of REML for these models 
bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)

#One model still won't converge
bad_models2 <- bad_mod_results %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(mods, by = c("metric" = "metric_name", "model_id"))

#Fit with a different algorithm
bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))

metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
          bad_mod_results %>% filter(is.na(warning)),
          bad_mod_results2 %>% filter(is.na(warning)))

usethis::use_data(metric_model_table)
pins::pin(metric_model_table, board = "github")
```

```{r}
# SES_rand_slope <- model_data %>%
#   filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
#   group_nest(metric) %>%
#   mutate(#year_center = map(data, ~scale(.x$year)),
#          model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
#                                  data = .x)),
#            # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
#            #                       data = .x)),
#            # test = map2(model, null_model, anova),
#            # test_res = map(test, tidy),
#            coef = map(model, tidy),
#            ) %>% 
#   select(metric, model, coef) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))
# 
# #check that they all used the same number of observations
# map(SES_rand_slope$model, nobs)
# 
# #these two metrics are cranky and want their own year scaling to converge
# 
# #Jaccard Base
# jacc_base_mod <-  model_data %>%
#   filter(metric == ("Jaccard_base")) %>%
#   mutate(year_scaled = scale(year)) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
#                                  data = .x)),
#            coef = map(model, tidy),
#            ) %>% 
#   select(metric, model, coef) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))
# 
# # S model
# s_mod <- model_data %>%
#   filter(metric %in%  c("S")) %>%
#   #it wants its own scaling to converge for some reason
#   mutate(year_scaled = scale(year)) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
#                                  data = .x, REML = TRUE)),
#            coef = map(model, tidy)
#            ) %>% 
#   select(metric, model, coef) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))
# 
# metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)
# 
# #save coefficients and models
# usethis::use_data(metric_model_table)
# pins::pin(metric_model_table, board = "github")
# ```
# Taxa model
# ```{r}
# taxa_rand_slope <- model_data %>%
#   filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
#   group_nest(metric) %>%
#   mutate(#year_center = map(data, ~scale(.x$year)),
#          model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
#                                  data = .x)),
#            # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
#            #                       data = .x)),
#            # test = map2(model, null_model, anova),
#            # test_res = map(test, tidy),
#            coef = map(model, tidy),
#            ) %>% 
#   select(metric, model, coef) %>%
#   unnest(cols = c(coef))
# 
# #Jaccard Base
# jacc_taxa_base_mod <-  model_data %>%
#   filter(metric == ("Jaccard_base")) %>%
#   mutate(year_scaled = scale(year)) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
#                                  data = .x)),
#            coef = map(model, tidy),
#            ) %>% 
#   select(metric, model, coef) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))
# 
#  s_taxa_mod <- model_data %>%
#   filter(metric %in%  c("S")) %>%
#   #it wants its own scaling to converge for some reason
#   mutate(year_scaled = scale(year)) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
#                                  data = .x, REML = TRUE)),
#            coef = map(model, tidy)
#            ) %>% 
#   select(metric, model, coef) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))
# 
# fric_taxa_mod <- model_data %>%
#   filter(metric %in%  c("SES_FRic")) %>%
#   #it wants its own scaling to converge for some reason
#   mutate(year_scaled = scale(year)) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
#                                  data = .x, REML = TRUE)),
#            coef = map(model, tidy)
#            ) %>% 
#   select(metric, model, coef) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))
# 
# taxa_model_table <- bind_rows(taxa_rand_slope, jacc_taxa_base_mod, s_taxa_mod, fric_taxa_mod)

@@

#### Analizing slopes
Is slope a function of timeseries characteristics?
```{r}
library(multcomp)

study_slope_data <- study_slopes %>%
  select(study_id, metric, estimate) %>%
  left_join(meta %>% 
              select(-rarefyid) %>% 
              distinct() %>%
              mutate(study_id = as.character(study_id))) %>%
  left_join(model_data %>% select(study_id, startyear, duration) %>% distinct())

taxa_slope <- study_slope_data %>%
  mutate(taxa = as.factor(taxa)) %>%
  group_nest(metric) %>%
  mutate(#year_center = map(data, ~scale(.x$year)),
         model = map(data, ~lmer(estimate ~ (1|study_id) + year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
           #                       data = .x)),
           # test = map2(model, null_model, anova),
           # test_res = map(test, tidy),
           coef = map(model, tidy),
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

#taxa model
taxa_slope <- study_slope_data %>%
  mutate(taxa = as.factor(taxa)) %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~glm(estimate ~ taxa, 
                                 data = .x)),
           tukey_mod = map(model, ~glht(., mcp(taxa="Tukey"))),
         tukey_coef = map(tukey_mod, tidy)) %>% 
  select(metric, tukey_coef) %>%
  unnest(cols = c(tukey_coef)) %>%
  rename(covariate = term)

climate_slope <- study_slope_data %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(#year_center = map(data, ~scale(.x$year)),
         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
         model = map(data, ~glm(estimate ~ climate, 
                                 data = .x)),
           tukey_mod = map(model, ~glht(., mcp(climate="Tukey"))),
         tukey_coef = map(tukey_mod, tidy)) %>% 
  select(metric, tukey_coef) %>%
  unnest(cols = c(tukey_coef)) %>%
  rename(covariate = term)

realm_slope <- study_slope_data %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(#year_center = map(data, ~scale(.x$year)),
         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
         model = map(data, ~glm(estimate ~ realm, 
                                 data = .x)),
           tukey_mod = map(model, ~glht(., mcp(realm="Tukey"))),
         tukey_coef = map(tukey_mod, tidy)) %>% 
  select(metric, tukey_coef) %>%
  unnest(cols = c(tukey_coef)) %>%
  rename(covariate = term)
